---
title: "class17"
format: gfm
editor: visual
---

# Getting started

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")

head(vax)
```

**Q1.** What column details the total number of people fully vaccinated?

persons_fully_vaccinated

**Q2.** What column details the Zip code tabulation area?

zip_code_tabulation_area

**Q3.** What is the earliest date in this dataset?

```{r}
min(vax$as_of_date)
```

**Q4.** What is the latest date in this dataset?

```{r}
max(vax$as_of_date)
```

```{r}
skimr::skim_without_charts(vax)
```

**Q5.** How many numeric columns are in this dataset?

14

**Q6.** Note that there are "missing values" in the dataset. How many `NA` values there in the `persons_fully_vaccinated` column?

```{r}
na_pfv <- sum(is.na(vax$persons_fully_vaccinated))
na_pfv
```

**Q7.** What percent of `persons_fully_vaccinated` values are missing (to 2 significant figures)?

```{r}
na_pfv/nrow(vax) * 100
```

**Q8.** \[Optional\]: Why might this data be missing?

It might be because the data hasn't been updated.

## Working with dates

```{r}
library(lubridate)
```

```{r}
today()
```

As the class of as_of_date column is character, we can't do math with it.

```{r}
class(vax$as_of_date)
```

```{r}
# This will give an error
# today() - vax$ax_of_date[1]
```

Convert our date date into a lubridate format things.

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

Now we can do math with dates.

```{r}
today() - vax$as_of_date[1]
```

Determine how many days the dataset span.

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

**Q9.** How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[nrow(vax)]
```

**Q10.** How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

# Working with ZIP codes

```{r}
library(zipcodeR)
```

```{r}
class(vax$zip_code_tabulation_area)
```

```{r}
geocode_zip('92037')
```

Calculate the distance between the centroids of two zip codes in miles.

```{r}
zip_distance('92037','92109')
```

Pull census data about ZIP code areas.

```{r}
reverse_zipcode(c('92037', '92109'))
```

**OPTIONAL:** use reverse_zipcode() to pull census data later on for any or all ZIP code areas we might be interested in.

```{r}
# Pull data for all ZIP codes in the dataset
# zipdata <- reverse_zipcode(vax$zip_code_tabulation_area)
```

# Focus on the San Diego area

Subset to San Diego county only areas.

```{r}
sd <- vax[vax$county == "San Diego", ]
```

Another way to do it.

```{r}
# library(dplyr)
# sd <- filter(vax, county == "San Diego")
```

```{r}
nrow(sd)
```

Using dplyr is often more convenient when we are subsetting across multiple criteria.

```{r}
# Subset to all San Diego county areas with a population of over 10000
# sd.10 <- filter(vax, county == "San Diego" & age5_plus_population > 10000)
```

**Q11.** How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

**Q12.** What San Diego County Zip code area has the largest population in this dataset?

```{r}
sd$zip_code_tabulation_area[which.max(sd$tot_population)]
```

**Q13.** What is the overall average (with 2 decimal numbers) "Percent of Population Fully Vaccinated" value for all San Diego "County" as of "2023-05-23"?

```{r}
library(dplyr)

sd_filter <- sd %>%
  filter(county == "San Diego", as_of_date == "2023-05-23")

mean(sd_filter$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

**Q14.** Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of "2023-05-23"?

```{r}
hist(sd_filter$percent_of_population_fully_vaccinated, xlab = "Percent of Population Fully Vaccinated on 2023-05-23", main = "Histogram of Vaccination Rates Across San Diego County")
```

## Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

**Q15**. Using **ggplot** make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
library(ggplot2)

ggplot(ucsd) +
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

ggplot

```{r}
library(ggplot2)

ggplot(sd_filter) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() +
  ggtitle("Histogram of Vaccination Rates Across San Diego County")
```

## Comparing to similar sized areas

```{r}
vax.36 <- filter(vax, age5_plus_population > 36144 & as_of_date == "2023-05-23")

head(vax.36)
```

**Q16**. Calculate the mean *"Percent of Population Fully Vaccinated"* for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* "2023-05-23". Add this as a straight horizontal line to your plot from above with the `geom_hline()` function?

```{r}
mean <- mean(vax.36$percent_of_population_fully_vaccinated)
mean
```

```{r}
ggplot(ucsd) +
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated") +
  geom_hline(yintercept = mean, col = 'red', linetype = 'dashed')
```

**Q17.** What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the *"Percent of Population Fully Vaccinated"* values for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* "2023-05-23"?

```{r}
summary(vax.36)
```

**Q18.** Using ggplot generate a histogram of this data.

```{r}
library(ggplot2)

ggplot(vax.36, na.rm=TRUE) +
  aes(x=percent_of_population_fully_vaccinated) +
  geom_histogram()
```

**Q19**. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax_92109 <- vax %>% filter(as_of_date == "2023-05-23") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)

vax_92040 <- vax %>% filter(as_of_date == "2023-05-23") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)

vax_92109$percent_of_population_fully_vaccinated > mean

vax_92040$percent_of_population_fully_vaccinated > mean
```

**Q20.** Finally make a time course plot of vaccination progress for all areas in the full dataset with a `age5_plus_population > 36144`.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
```

```{r}
ggplot(vax.36.all) +
  aes(as_of_date, percent_of_population_fully_vaccinated, group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color='blue') +
  ylim(c(0,1)) +
  labs(x='Date', y='Percent Vaccinated', title="Vaccination rate accross California", subtitle="Only areas with a population above 36k are shown.") +
  geom_hline(yintercept=mean, linetype='dashed')
```
