---
title: "class10_project"
author: "Xueran Zou"
date: 2023-05-05
format: gfm
editor: visual
---

# Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names = 1)
head(candy)
```

**Q1**. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

**Q2**. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity == 1)
```

# What is your favorite candy?

Fine the winpercent value for Twix

```{r}
candy["Twix", ]$winpercent
```

**Q3**. What is your favorite candy in the dataset and what is it's `winpercent` value?

```{r}
candy["Air Heads", ]$winpercent
```

**Q4**. What is the `winpercent` value for "Kit Kat"?

```{r}
candy["Kit Kat", ]$winpercent
```

**Q5**. What is the `winpercent` value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

**Q6**. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Sugarpercent, pricepercent, winpercent.

**Q7**. What do you think a zero and one represent for the `candy$chocolate` column?

A zero represents that the candy contains chocolate while the one represents that the candy contains no chocolate.

**Q8**. Plot a histogram of `winpercent` values

```{r}
library(ggplot2)
ggplot(candy, aes(x = winpercent)) + geom_histogram(binwidth = 10)

hist(candy$winpercent)
```

**Q9**. Is the distribution of `winpercent` values symmetrical?

No.

**Q10**. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
```

It's below 50%.

**Q11**. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean_chocolate <- mean(candy$winpercent[as.logical(candy$chocolate)])
mean_fruity <- mean(candy$winpercent[as.logical(candy$fruity)])
mean_chocolate > mean_fruity
```

**Q12**. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

```{r}
boxplot(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

# Overall Candy Rankings

```{r}
library(dplyr)
```

**Q13**. What are the five least liked candy types in this set?

```{r}
candy %>%
  arrange(desc(winpercent))
```

**Q14**. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>%
  arrange(winpercent)
```

**Q15**. Make a first barplot of candy ranking based on `winpercent` values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent,rownames(candy)) +
  geom_col()
```

**Q16**. This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent) )+
  geom_col()
```

Add some colors.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

**Q17**. What is the worst ranked chocolate candy?

Nik L Nip

**Q18**. What is the best ranked fruity candy?

Reeses Peanut Buffer cup

# Taking a look at pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols, size=3.3, max.overlaps=5)
```

**Q19**. Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

```{r}
candy %>%
  arrange(desc(winpercent)) %>%
  head(5)
```

**Q20**. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

**Q21**. Make a barplot again with `geom_col()` this time using `pricepercent` and then improve this step by step, first ordering the x-axis by value and finally making a so called "dot chat" or "lollipop" chart by swapping `geom_col()` for `geom_point()` + `geom_segment()`.

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()
```

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

# Exploring the correlation structure

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

**Q22**. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolat

**Q23**. Similarly, what two variables are most positively correlated?

Winpercent and chocolate, chocolate and bar.

# Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)

ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

**Q24**. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard and pluribus.
